---
title: "Example Analysis"
---

This example analysis explores the SEER Breast Cancer Dataset from the National Cancer Institute (NCI).

The question I aim to answer is: how does the race of the patient affect their status, survival, and other features?

::: callout-note
While I acknowledge that we would need a deeper analysis, such as perhaps a comparison of genomic data (e.g., BRCA 1/2 status), I thought this would be an interesting way to look at how race and status are related in this dataset.
:::

### Installing packages and loading libraries

```{r loading the libraries we will use}
#| output: false
library(tidyverse)
```

### Exploring the dataset to be used

```{r reading in the data}
#| output: false
bc <- read_csv("data/SEER_Breast_Cancer_Dataset.csv")
```

```{r making the data a tibble and renaming columns}
bc <- bc %>%
  rename(age = Age,
         race = Race,
         marital_status = "Marital Status",
         t_stage = "T Stage",
         n_stage = "N Stage",
         sixth_stage = "6th Stage",
         grade = "Grade",
         a_stage = "A Stage",
         tumor_size = "Tumor Size",
         estrogen_status = "Estrogen Status",
         progesterone_status = "Progesterone Status",
         regional_node_examined = "Regional Node Examined",
         regional_node_pos = "Reginol Node Positive",
         survival_months = "Survival Months",
         status = Status) %>%
  mutate_if(is.character, as.factor)
```

#### Data Dictionary

| Variable               | Class  | Description                                                                                                |
|------------|------------|------------------------------------------------|
| age                    | double | age at diagnosis                                                                                           |
| race                   | factor | race recode (white, black, other)                                                                          |
| marital_status         | factor | marital status at diagnosis                                                                                |
| t_stage                | factor | Breast Adjusted AJCC 6th T (1988+)                                                                         |
| n_stage                | factor | Breast Adjusted AJCC 6th N (1988+)                                                                         |
| sixth_stage            | factor | Breast Adjusted AJCC 6th Stage (1988+)                                                                     |
| grade                  | factor | grade as defined in ICD-O-2; 1992                                                                          |
| a_stage                | factor | SEER historic stage A                                                                                      |
| tumor_size             | double | CS tumor size (2004+)                                                                                      |
| estrogen_status        | factor | ER Status Recode Breast Cancer (1990+)                                                                     |
| progesterone_status    | factor | PR Status Recode Breast Cancer (1990+)                                                                     |
| regional_node_examined | double | total number of regional lymph nodes that were removed and examined by the pathologist                     |
| regional_node_pos      | double | the exact number of regional lymph nodes examined by the pathologist that were found to contain metastases |
| survival_months        | double | survival months                                                                                            |
| status                 | factor | vital status recode (1 = alive, 2 = dead)                                                                  |

: Data dictionary for the SEER Breast Cancer Dataset

### Building our plots/visualization

First, I want to see the total number of patients that are either alive or dead by race.

```{r see total number alive/dead by race}
bc_status <- bc %>%
  group_by(race, status) %>%
  summarise(total = n()) %>%
  ggplot(aes(fill=status, y=total, x=race)) +
    geom_bar(position="dodge", stat="identity") +
    scale_x_discrete(labels=c('Black', 'Other (not Black/White)', 'White'))

bc_status
```
